%--------------------------------------------------------------------------------------------------
% 
\chapter{Introduction}
%--------------------------------------------------------------------------------------------------

\section{Background and Motivation}

Flooding is a major natural hazard, responsible for substantial economic losses, social disruption, and loss of life across the globe. With increasing urbanization and changing climate conditions, flood risks are intensifying in both frequency and magnitude. The Intergovernmental Panel on Climate Change (IPCC) projects a rising likelihood of extreme hydrological events, including both pluvial and fluvial floods, due to higher precipitation variability and rising sea levels.

In Europe, and particularly in Central and Eastern regions such as Slovenia, recent flood events have exposed vulnerabilities in both infrastructure and institutional response systems. The need for robust flood risk models that can inform spatial planning, insurance pricing, and asset management is therefore more pressing than ever. This is especially relevant for stakeholders in the real estate and financial sectors, who must now account for physical climate risks as part of their environmental, social, and governance (ESG) reporting and compliance requirements.

\section{Climate Risk and the ESG Framework}

Climate-related physical risks, such as floods, heatwaves, and storms, are becoming a critical component of ESG disclosures. Investors and financial institutions are increasingly required to assess and report the exposure of their assets to climate hazards. In this context, the concept of physical risk is typically decomposed into three components:

\begin{itemize}
    \item \textbf{Hazard} – the probability and intensity of the climate-related event (e.g., flood depth, wind speed).
    \item \textbf{Vulnerability} – the expected damage to an asset given the intensity of the hazard.
    \item \textbf{Exposure} – the value or presence of assets in harm’s way.
\end{itemize}

The combination of these three factors forms the basis of quantitative risk models that estimate potential economic losses due to climate events. Yet, despite the wide adoption of such frameworks, the underlying models are often poorly validated, especially at local or parcel-level scales relevant for property-level decision making.

\section{Problem Statement}

While global flood hazard models such as the WRI Aqueduct River Floods provide broad coverage and climate scenario integration, they are often limited in spatial resolution and may not reflect local hydrological conditions. On the other hand, high-resolution local models, such as Slovenia’s \textit{Integralna karta globin} (IKG), offer detailed inundation mapping but are typically fragmented, static (i.e., not climate scenario-aware), and difficult to integrate into broader risk platforms.

Furthermore, flood vulnerability is typically modeled using generalized depth--damage functions derived from empirical or synthetic data. These functions may perform well in aggregate, but their applicability across different building types, geographies, and damage-reporting standards remains uncertain. This creates uncertainty in downstream financial estimates and limits the usefulness of such models for asset-level ESG disclosure.

\section{Research Objectives}

This thesis aims to evaluate the performance of global and local flood risk models with a specific focus on hazard and vulnerability components. The main objectives of the research are:

\begin{enumerate}
    \item To compare the predictive accuracy of flood depth estimates from a global model (WRI Aqueduct) and a high-resolution Slovenian model (IKG) using real flood observations.
    \item To assess how well global depth--damage functions (e.g., from the JRC) reflect reported flood damage across different real estate asset types in Slovenia.
    \item To evaluate whether spatial resolution or proximity to water bodies significantly influence model performance.
    \item To develop and document a data onboarding pipeline that enables integration of heterogeneous local flood maps into the modular \textit{Physrisk} physical risk modeling framework.
\end{enumerate}

\section{Methodological Approach}

The thesis adopts a comparative modeling approach using geospatial and statistical techniques. A validation dataset, compiled from the records of the Slovenian Administration for Civil Protection and Disaster Relief, includes appraised damage and measured flood depths for selected events between 2010 and 2023.

Flood hazard models are evaluated using both regression-based and classification-based metrics to assess depth prediction accuracy. Vulnerability models are tested by comparing observed versus predicted damages using segmented analysis by flood depth and real estate type. All data are processed and harmonized into a common geospatial reference system and stored in Zarr format, facilitating efficient access within the \textit{Physrisk} platform.

\section{Scientific and Practical Relevance}

From a scientific perspective, this work contributes to the ongoing effort of validating global hazard and vulnerability models in local contexts. By empirically assessing model accuracy, the thesis provides critical insight into where global models fall short and where local models offer meaningful advantages.

Practically, the work supports the advancement of data-driven ESG disclosures by enhancing the interoperability and transparency of physical risk assessment tools. The technical contributions also provide a template for integrating fragmented local hazard datasets into scalable, modular frameworks like \textit{Physrisk}, which are increasingly used in institutional risk modeling and regulatory reporting.

\section{Structure of the Thesis}

The remainder of the thesis is structured as follows:

\begin{itemize}
    \item \textbf{Chapter 2} presents a review of relevant literature and related work on flood hazard modeling, vulnerability estimation, and ESG-aligned risk frameworks.
    \item \textbf{Chapter 3} describes the study area, datasets, and the methodological framework used for model evaluation and data processing.
    \item \textbf{Chapter 4} details the integration of the local IKG model into the Physrisk framework and presents the architecture of the comparison toolset.
    \item \textbf{Chapter 5} provides experimental results, including quantitative evaluation of model performance, depth–damage function analysis, and financial impact estimates.
    \item \textbf{Chapter 6} discusses the key findings, limitations, and possible future research directions, and concludes the thesis.
\end{itemize}

% \section{Thesis Structure}

% The thesis should be structured as follows (note that while some parts are optional, their order is not):
% \begin{itemize}
% 	\item \emph{Front Matter}
% 	\begin{itemize}
% 		\item Title pages
% 		\item Dedication (optional)
% 		\item Acknowledgments
% 		\item Abstract
% 		\item Povzetek
% 		\item Contents
% 		\item List of Figures (required if the thesis contains figures)
% 		\item List of Tables (required if the thesis contains tables)
% 		\item List of Algorithms (required if the thesis contains algorithms)
% 		\item Abbreviations (optional)
% 		\item Symbols (optional)
% 		\item Glossary (optional)
% 	\end{itemize}
% 	\item \emph{Main Matter}
% 	\begin{itemize}
% 		\item 1 Introduction
% 		\item $2, \dots, n-1$ Chapters
% 		\item $n$ Conclusions
% 	\end{itemize}
% 	The Introduction must clearly summarize the thesis from the topic application of the master thesis. The core text of the master thesis can be substituted by publications (or papers accepted for publication) in internationally recognized journals. In this case, the Introduction should clearly describe the scientific method and the candidate’s contribution to any publication which has been produced by several authors. In the Discussion or Conclusions, the candidate should summarize coherently the results of his/her thesis.
% 	\item \emph{Appendices} (optional)
% 	\begin{itemize}
% 		\item Appendix A Title
% 		\item Appendix B Title
% 		\item \dots
% 	\end{itemize}
% 	If desired, the candidate can place some of his/her relevant papers in the appendices. In order to do so, he/she must obtain permissions from publishers.
% 	\item \emph{Back Matter}
% 	\begin{itemize}
% 		\item References
% 		\item Bibliography 
% 		\item Biography
% 		\item Index (optional)
% 	\end{itemize}
% \end{itemize}

% In addition, each author must prepare the cover page and spine using the two \texttt{.doc} files provided by IPS.

% \section{Sectioning Example}

% This is a section. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subsection{Subsection}

% This is a subsection. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subsubsection{Subsubsection}

% This is a subsubsection. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \paragraph{Paragraph}

% This is a paragraph. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text.  

% Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subparagraph{Subparagraph}

% This is a subparagraph. It is not possible to get deeper than this. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subparagraph{Subparagraph}

% This is a subparagraph. It is not possible to get deeper than this. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \paragraph{Paragraph}

% This is a paragraph. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subsubsection{Subsubsection}

% This is a subsubsection. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text.  

% \paragraph{Paragraph}

% This is a paragraph. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text.  

% \paragraph{Paragraph}

% This is a paragraph. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subsection{Subsection}

% This is a subsection. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subsubsection{Subsubsection}

% This is a subsubsection. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 

% \subsubsection{Subsubsection}

% This is a subsubsection. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. Some more text. 